---
title: "Dataquest: Guided Project: Investigating COVID-19 Virus Trends"
output: html_document
authour: Ba Linh Le
---
# Introduction: 

This is a guided project provided by Dataquest's course "Introduction to R. Data Structures in R". To make it a bit more interesting, I have attempted both a dplyr and a base R solution. The following project analyzes demographic cross-country data on COVID-19. The dataset is from Kaggle and available for download here: https://www.kaggle.com/lin0li/covid19testing 

The analysis will provide an answer to the question which countries have had the highest number of positive cases against the number of tests. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tibble)
library(dplyr)
```

```{r data exploration}
## Import dataset
covid_df <- read.csv("covid19.csv")

## Number of rows and columns
dim(covid_df)

## Names of the columns 
vector_cols <- colnames(covid_df)
vector_cols

## Peek on dataset snippets
head(covid_df)
glimpse(covid_df)
```

```{r base R approach}
## Extracting country-level data
covid_df_all_states <- covid_df[covid_df$Province_State == "All States", !(colnames(covid_df)) == "Province_State"]

## Filtering columns related to daily measures
covid_df_all_states_daily <- covid_df_all_states[c("Date", "Country_Region", "active", "hospitalizedCurr", "daily_tested", "daily_positive")]
``` 


```{r }
# Question: What is the overall number of COVID-19 tested, positive, active and hospitalized cases by country? What are the top tend countries? 

## Using aggregate() to sum daily information by country
covid_df_all_states_daily_sum <- aggregate(cbind(daily_tested, daily_positive, active, hospitalizedCurr) ~ Country_Region, data = covid_df_all_states_daily, sum, na.rm = TRUE)

## Arrange data by the sum of daily tests and extract first ten rows
covid_df_all_states_daily_sum <- covid_df_all_states_daily_sum[order(covid_df_all_states_daily_sum$daily_tested, decreasing = TRUE),]
covid_top_10 <- covid_df_all_states_daily_sum[1:10,]
covid_top_10
```

```{r }
# Question: Which countries have had the highest number of positive cases against the number of tests? 

## Assign the estimates to new vectors
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$daily_tested
positive_cases <- covid_top_10$daily_positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalizedCurr

## Use sapply() to assign the character vector countries to all new vectors
x <- sapply(list(tested_cases, positive_cases, active_cases, hospitalized_cases), setNames, countries)

## Divide the vector positive cases by the vector tested cases
positive_tested_top_3 <- x[,2]/x[,1]

## Arrange by the ratio 
positive_tested_top_3 <- positive_tested_top_3[order(positive_tested_top_3, decreasing = TRUE)]

## Extract the top three tows
positive_tested_top_3 <- positive_tested_top_3[1:3]
positive_tested_top_3
```

```{r }
# Task: Create a matrix to store all the information available for the top 3 countries that have had the highest number of positive cases against the number of tests.

## Manually create the vectors 
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

## Combine the vectors in a matrix
covid_mat <- rbind(united_kingdom, united_states, turkey)

## Assign columnnames
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")
covid_mat
```

```{r }
# Task: Create a list to store all results so far. 

## Store question and answer manually 
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)

## Create the list to store data frames, matrix and vectors 
data_structure_list <- list(
  "dataframe" = list(covid_df, covid_df_all_states, covid_df_all_states_daily, covid_top_10), 
  "matrix" = list(covid_mat),
  "vector" = list(vector_cols, countries)
)

## Create the list to combine previous created list with question and answer
covid_analysis_list <- list(question, answer, data_structure_list)

## Display second element of the list
covid_analysis_list[[2]]
```


```{r dplyr approach}
## Data wrangling
covid_df_all_states <- covid_df %>%
  filter(Province_State == "All States") %>% 
  select(-Province_State)

covid_df_all_states_daily <- covid_df_all_states %>% 
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)

# Question: What is the overall number of COVID-19 tested, positive, active and hospitalized cases by country? What are the top tend countries? 
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>% 
  group_by(Country_Region) %>% 
  summarise(tested = sum(daily_tested),
            positive = sum(daily_positive),
            active = sum(active),
            hospitalized = sum(hospitalizedCurr)) %>% 
  arrange(desc(tested))

covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_top_10

# Question: Which countries have had the highest number of positive cases against the number of tests? 
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized

names(tested_cases) <- countries
names(positive_cases) <- countries
names(hospitalized_cases) <- countries
names(active_cases) <-  countries

positive_tested_top_3 <- data.frame(ratio = positive_cases/tested_cases) %>% 
  arrange(desc(ratio)) %>% 
  slice(1:3)

# Task: Create a matrix to store all the information available for the top 3 countries that have had the highest number of positive cases against the number of tests.
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)
covid_mat <- rbind(united_kingdom, united_states, turkey)
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")
covid_mat

# Task: Create a list to store all results so far. 
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)
data_structure_list <- list(
  "dataframe" = list(covid_df, covid_df_all_states, covid_df_all_states_daily, covid_top_10), 
  "matrix" = list(covid_mat),
  "vector" = list(vector_cols, countries)
)

covid_analysis_list <- list(question, answer, data_structure_list)
covid_analysis_list[[2]]
```