---
title: "Dataquest: Guided Project: Creating An Efficient Data Analysis Workflow, Part 2"
output: html_document
author: Ba Linh Le 
---
# Introduction 

The aim of this project is to figure out if the company's new program helped to improve sales. The data is available here: https://data.world/dataquest/book-sales-data

```{r setup, message = FALSE}
library(readr)
library(stringr)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```


``` {r}
dat <- read_csv("sales2019.csv")
```
# Exploration

```{r}
dim(dat)
sapply(dat, class)
colSums(is.na(dat))
```

1. The dataset contains 5000 rows and 5 columns. The column names are: date, user_submitted_review, title, total_purchased, customer_type. 

2. All columns except for total_purchased are character vectors. Total_purchased is of the numeric type. 

3. There is missing data in user_submitted_review (885) and total_purchased (718). 

# Data preparation

## Manage NA cases 
```{r}
# Remove NA in user_submitted_review 
dat1 <- dat %>% 
  filter(!is.na(user_submitted_review))

# Calculate average number of books purchased
mean_purchase <-summarize(dat1, mean(total_purchased, 
                                     na.rm = TRUE))

# Fill missing values in total_purchased with average number of books purchased 
dat2 <- dat1 %>% 
  mutate(
    total_p = unlist(ifelse(is.na(total_purchased),
                      round(mean_purchase,2), 
                      total_purchased)))
```

## Classify text data 
```{r}
# Examine unique cases in user_submitted_review
unique(dat2$user_submitted_review)

# Function that takes in a sentence and returns a value indicating direction and degree of feedback 

classify_review <- function(x) {
  positive_review = case_when( 
    str_detect(x, "Awesome") ~ TRUE,
    str_detect(x, "lot") ~ TRUE,
    str_detect(x, "Never") ~ TRUE,
    str_detect(x, "Ok") ~ TRUE,
    TRUE ~ FALSE
  )
}

dat3 <- dat2 %>% 
  mutate(positive = classify_review(user_submitted_review))
```

# Analysis 

## Was the new book program effective in increasing book sales?
```{r}
# Conversion of date column to temporal data 
dat3$date <- mdy(dat3$date)

# Create grouping column to distinguish between sales before and after July 1 2019 
dat4 <- dat3 %>% 
  mutate(
    program = case_when( 
      date > "2019/07/01" ~ "After", 
      date < "2019/07/01" ~ "Before") 
  )

# Create summary table to compare number of books purchased before July 1, 2019 to after 
dat4 %>% group_by(program) %>% 
  summarize(total_books = sum(total_p))
```

The simple comparison shows that there is no substantial differences in book sales prior to and after the program implementation.

## Was the new book program effective in increasing book sales for subgroups?
```{r}
# Create summary table grouped by date and customer_type 
dat4 %>% group_by(program, customer_type) %>% 
  summarize(total_books = sum(total_p))
```

There is a 100 books difference for the book sales of both customer types. Businesses purchased around 100 books more, whereas individuals purchased around 100 books less.

``` {r, warning = FALSE}
# Create summary table grouped by date and customer_type 
dat4 %>% group_by(program, title) %>% 
  summarize(total_books = sum(total_p)) %>% 
  pivot_wider(., names_from = program,
              values_from = total_books) %>% 
  mutate(difference = After - Before) %>% 
  arrange(desc(abs(difference)))
```

The book sales increased and decreased for some book titles. The biggest differences were reported by "Fundamentals of R For Beginners" (decrease by ca. 265 books) and the "Secrets of R for Advanced Students" (increase by ca. 182 books).

## Did review scores improve as a result of the program?
```{r}
# Create summary table to compare number of positive reviews before and after July 1, 2019 
dat4 %>% group_by(program) %>% 
  summarize(n = sum(positive))
```

Book review scores improved slightly as a result of the program.

## Is there any relationship between month and the amount of books that were sold?
```{r}
# Classify months 
dat5 <- dat4 %>% 
  mutate(
    month = case_when(
      substr(dat4$date, 6, 7) == "01" ~ "January",
      substr(dat4$date, 6, 7) == "02" ~ "February",
      substr(dat4$date, 6, 7) == "03" ~ "March",
      substr(dat4$date, 6, 7) == "04" ~ "April",
      substr(dat4$date, 6, 7) == "05" ~ "May",
      substr(dat4$date, 6, 7) == "06" ~ "June",
      substr(dat4$date, 6, 7) == "07" ~ "July",
      substr(dat4$date, 6, 7) == "08" ~ "August",
      substr(dat4$date, 6, 7) == "09" ~ "September",
      substr(dat4$date, 6, 7) == "10" ~ "October",
      substr(dat4$date, 6, 7) == "11" ~ "November",
      substr(dat4$date, 6, 7) == "12" ~ "December",
      
    )
  )

# Create summary table to compare book sales per month
dat5 %>% group_by(month) %>% 
  summarize(sum = sum(total_p)) %>% 
  arrange(desc(sum))

# Scatterplot 
dat5$month <- factor(dat5$month, levels = month.name)
ggplot(aes(x = month), data = dat5) +
  theme_minimal() +
  geom_bar() + 
  xlab("Date") + 
  ylab("Amount of book orders") + 
  ggtitle("Bar chart")
```

Book sales appear to have decreased after the introduction of the program in July but increased again in December.